---
title: "Temporal evolution of notoriety of Wikipedia pages with origin in social networks"
author: "JJ Merelo and Elena Merelo"
date: "28 de febrero de 2017"
output:
  pdf_document:
    keep_tex: true
bibliography: refs.bib
abstract:  "The Wikipedia is a web portal created by users and its
  simplicity, references and also the inclusion as insets 
  introductory paragraphs for their pages in Google search results have made it the go-to place
  to find out about current events or persons featured in them. Besides, its open application programming interface (API)
  allows any user to know about the number of visits some particular
  page has. In this paper, after certain events that made Copernicus a
  viral meme in Spain, we study the intensity and duration of the increment of visits to his
  page and other pages related to the event. Using pages related to other persons
  as a comparison, we try to establish a typical duration of notoriety achieved through social networks,
  mainly comparing with visits in previous years in the same dates. We
  conclude that a 7 day duration is the statistical mode and that this duration is relatively
  independent of the initial increase in the number of visits."
---

# Introduction

The so called “Slashdot Effect”
[@adler1999slashdot,@halavais2001slashdot], known since the end of the
90s, is the name attributed to the increase in pageviews that the
websites experience when mentioned by publications
like [Slashdot](http://slashdot.org), one of the most visited in the
area of technology and especially programming. This effect provokes a
peak in visits, especially directed to a single webpage, that in a lot
of cases and more obscure ages in which there were not any publication
of content networks, had a similar effect of a denial of service
attack. In Spain we have also referred to the “Efecto Menéame”
[@dans2008anatomia], because of the influence of this aggregator of
news and even, before and in a lesser measure, the "Efecto Barrapunto"
[@blanco2009blog] originated by mentions to the web Barrapunto.com, a
Spanish Slashdot. 

However and to the extent of our knowledge, until now there has not
been many studies analising a similar effect aroused by mentions in
social networks, maybe due to the fact that such a effect is not
caused by a single website and might be eventually directed to many
different pages; even lesser is the effect they have in the search or
mindshare of the concepts. Indeed, web analytics is currently far more
complex and generally the visits on a page, say, of a newspaper or
ecommerce site are not public. And besides, instead of studying the
effect of a single link, nowadays it is of more interest to study the
topics that pop up in the social networks. However, the effect these
have in the mindshare of the public and its behaviour while surfing
the Net has not received so much attention as the behavior of the
trending topics themselves. 

At the start of 2017, an anecdote in a Wise Men from Orient parade in
Madrid, in which a video locutor referred to a character in a float as
Colon, to which he answered “I am Copernico!” in an amusing
tone,
[as is narrated in Verne](http://verne.elpais.com/verne/2017/01/07/articulo/1483785572_720611.html),
gave way to a storm of memes, which increased on the following days
when a journalist and a politician discussed in a TV programme about
what did Copernico really do. 

In a series of articles and figures written in Spanish [merelo17, @merelo17:2, @Merelo2017:figshare]
that show the temporary evolution of this mindshare through the visits
to the Wikipedia page of Copernico [@wiki:Copernico] we have analyzed
the change in the number of visits, especially looking for the
duration of this phenomenon by means of studying the daily
number of visits. In this article, that closes the series, we try to find the causes of that duration, comparing it with
other phenomenons of visits to the Wikipedia which have happenned in a
similar time frame.

The rest of the paper is organised in the following way: first we present
the methodology that has been followed in order to capture and study
the data is presented. In the following section the visits to the
Wikipedia page of Copernico and other
related webpages is analized, along with the evolution in its
ranking of viewed pages in the Wikipedia. We will also examine the
visits to the pages of other persons that have become notorious due to
several reasons. Finally, we will draw some conclusions. 

## Methodology


The Wikipedia, through its application programming interface (API),
allows acces to the visits of each and every one of the different
pages it hosts. It is a REST interface which can be directly accessed
without any authentication. For the purposes of this paper, we tap
this API using  a script that, along with the rest of data and
programmes, is free software and hence can be downloaded from GitHub
[@Merelo2017:figshare]. The script uses `curl` for downloading the
file, just by codifying as URL the name of the page, next to
`all-agents`, and selecting as parameters the two dates in which we
are interested, from the start of 2017 until now. Due  to the fact
that the Wikipedia API returns a file in JSON, we pull the data we are
interested 
in out by means of the `jq` utility, that allows to make complex
queries to JSON files. That way, we extract the views, and we put it
in data files, which will also be attached to this paper. The rest of
the data analyzed in this paper is extracted in the same way, via
shell scripts that download an URL and the use of `jq`. 

 Nonetheless, in order to extract the ranking of the one thousand most
 visited pages, a script that gets the position in the ranking and the
 number of visits, along with the title of the article, is used. Then,
 the information is filtered, leaving only the "content" articles that are in
 the ranking, deleting special pages. Furthermore the ranking is once
 again calculated. Last, these rankings are processed to calculate the
 evolution in the position of the page as the days go by, saving in a
 file a number with the position that a specific page occupies each
 day. 
 
The results of analyzing the data are shown in the next section.


## Visits to the page of Nicolás Copérnico and other related pages

Data is processed in this paper using R; scripts can be accessed in
the source to this paper.

```{r setup,echo=FALSE}
library(ggplot2)
library(ggthemes)
datos.2016 <- read.table('datos/Nicol%C3%A1s_Cop%C3%A9rnico-2016.dat')
datos.2017 <- read.table('datos/Nicol%C3%A1s_Cop%C3%A9rnico-2017.dat')
datos <- data.frame( dia=c(1:nrow(datos.2016),1:nrow(datos.2017)),
                    year=c(rep('2016',nrow(datos.2016)),rep('2017',nrow(datos.2017))),
                    visitas=c(datos.2016$V1,datos.2017$V1))
```               

First we will focus on the anecdote that originated this paper by
analyzing the visits to the Nicolás Copérnico page in the Spanish
wikipedia during 2017 and the same period in 2016. This period falls
outside the school year in Spain, with classes starting just next to
that day. That might account for a certain interest in the page, which
implies that there will be some daily variation in the number of
visits anyway. Comparing the visits for two years will help us
discount that effect. 

```{r imagen,echo=FALSE}
ggplot(datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()
```
It is noticeable that, after reaching a peak the first day, it
descends the second one and even quicklier the third
one. Nevertheless, from the fourth day the fall is much softer, until
having approximately the same data a week later. In fact, there is
some variation also in the previous year, that is why we compare  the
relationship between last and this year´s visits in the next chart.

```{r ratio,echo=FALSE}
ratio <- data.frame( dia=c(1:nrow(datos.2016)),ratio=datos.2017$V1/datos.2016$V1)
ggplot(ratio,aes(x=dia,y=ratio))+geom_line()+theme_tufte()
```
From an initial situation where the number of visits is similar,
though slightly superior, it goes to having 20 times more visits,
number that slowly drops to 10 times three days past the peak. On the
fifth day the rate is of a 50% more visits; a little bit over last
year´s on the same date, taking into account that last year´s tendency
was of a gradual growth in the number of visits. Eventually, the
number of visits reaches a plateau. 

It is interesting to have a look at the evolution of some pages linked
to the same concept, especially the one of the main achievement of
Copérnico, the heliocentric theory, which motivated the following day
a debate on TV. 

```{r helio,echo=FALSE}
hc.2016 <- read.table('datos/Teor%C3%ADa_helioc%C3%A9ntrica-2016.dat')
hc.2017 <- read.table('datos/Teor%C3%ADa_helioc%C3%A9ntrica-2017.dat')
hc.datos <- data.frame( dia=c(1:nrow(hc.2016),1:nrow(hc.2017)),
                    year=c(rep('2016',nrow(hc.2016)),rep('2017',nrow(hc.2017))),
                    visitas=c(hc.2016$V1,hc.2017$V1))
ggplot(hc.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()
```
The number of visits moves in a different scale and thus there are
more fluctiations along the day. However, like in the previous case,
even though initially the visits were very much alike or even
inferiors, presumably in five days it went back to its initial value,
with a small variation. In five days it has less visits than last
year, the same situation that the previous day to the incident. 

What also changes is the correlation between the number of people that
visit a web and the ones that visit the other. We have a look at it in the next graphic.

```{r ratiohc,echo=FALSE}
ratio.hc.2016 <- datos.2016/hc.2016
ratio.hc.2017 <- datos.2017/hc.2017
hc.ratios <- data.frame( dia=c(1:nrow(ratio.hc.2016),1:nrow(ratio.hc.2017)),
                    year=c(rep('2016',nrow(ratio.hc.2016)),rep('2017',nrow(ratio.hc.2017))),
                    visitas=c(ratio.hc.2016$V1,ratio.hc.2017$V1))
ggplot(hc.ratios,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()
```

In the year 2016 there is a rate inferior to 2, consequently the page
of the heliocentric theory receives more or less half the visits that
Copérnico´s, with little variations. We suppose that there is more
interest in the character than in its contribution, though not a
lot. On the other hand, when the issue “Copérnico-Ojeda" breaks out,
all of a sudden the people get curious about the character, without
caring much about his contribution to the scientific world. It is to
important to point out that we do not know whether they are the same
people that visit the webpage, because of the fact that the path they
follow in Wikipedia is not returned by the API. Anyway, it is reasonable
to suppose that a big percentage of the people that visit
whichever of the pages will end up following the links in order to
visit the other one. The link to the heliocentric theory appears in
the first lines of Copérnico´s page and it is the thirteenth link from
the start of the page, while Copérnico shows up int the second
paragraph of the theory page, being the tenth link in this case.

Let´s also see another page, Galileo Galilei´s, who spread the
heliocentric theory. His name  does not show up in any way in the
incident but to check the visits here would be relevant to study the interest these kind of
pages have and whether there is a offshoot of interest from one to
another. 

```{r galileo,echo=FALSE}
galileo.2016 <- read.table('datos/2016-galileo.dat')
galileo.2017 <- read.table('datos/2017-galileo.dat')
galileo.datos <- data.frame( dia=c(1:nrow(galileo.2016),1:nrow(galileo.2017)),
                    year=c(rep('2016',nrow(galileo.2016)),rep('2017',nrow(galileo.2017))),
                    visitas=c(galileo.2016$V1,galileo.2017$V1))
ggplot(galileo.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()
```

Apart from the curiosity on  do they evolve in the same way, with a
considerable drop the seventh day, there is neither an obvious
tendency nor can we suppose that pages of this kind of content could
have an increased interest. In fact, the twelfth day there were less
visits in Galileo´s page than last year. Although there is a chance of
a transfer of attention and hence of visits from one to another, since
Galileo´s is linked to Copérnicos, and as there aren´t changes in the
same way as in Copérnico´s rules out a sudden interest in the
Renaissance Science and the heliocentric theory like a cause of the
rise in visits that will be then attributable to the incident. 

From this analysis we can conclude that the interest in a particular
Wikipedia page famous through a meme in the social networks lasts
around seven days, and that it is a shallow interest extending
exclusively to the page itself, not any concepts related to it. Let us
see next what happens to other similar pages.

## Duration of notability in other Wikipedia pages. 

We also examine the visits to the page of Meryl Streep, whose speech during the Golden Globes had a considerable impact, in the same way as the elected president´s reaction did. In this case it is an example of a page that on its own has a lot of visits:

```{r meryl,echo=FALSE}
meryl.2016 <- read.table('datos/Meryl_Streep-2016.dat')
meryl.2017 <- read.table('datos/Meryl_Streep-2017.dat')
john.2016 <- read.table('datos/John_Hurt-2016.dat')
john.2017 <- read.table('datos/John_Hurt-2017.dat')
meryl.datos <- data.frame( dia=c(1:nrow(meryl.2016),1:nrow(meryl.2017)),
                    year=c(rep('2016',nrow(meryl.2016)),rep('2017',nrow(meryl.2017))),
                    visitas=c(meryl.2016$V1,meryl.2017$V1))
ggplot(meryl.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_point()+geom_smooth(span=0.25)+theme_tufte()
john.datos <- data.frame( dia=c(1:nrow(john.2016),1:nrow(john.2017)),
                    year=c(rep('2016',nrow(john.2016)),rep('2017',nrow(john.2017))),
                    visitas=c(john.2016$V1,john.2017$V1))
ggplot(john.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()
carrie.2016 <- read.table('datos/Carrie-Fisher-2016.dat')
carrie.2017 <- read.table('datos/Carrie-Fisher-2017.dat')
carrie.datos <- data.frame( dia=c(1:nrow(carrie.2016),1:nrow(carrie.2017)),
                    year=c(rep('2016',nrow(carrie.2016)),rep('2017',nrow(carrie.2017))),
                    visitas=c(carrie.2016$V1,carrie.2017$V1))
ggplot(carrie.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_point()+geom_smooth(span=0.25)+theme_tufte()

# George Michael
george.2016 <- read.table('datos/George--2016.dat')
george.2017 <- read.table('datos/George-Michael-2017.dat')
george.datos <- data.frame( dia=c(1:nrow(george.2016),1:nrow(george.2017)),
                    year=c(rep('2016',nrow(george.2016)),rep('2017',nrow(george.2017))),
                    visitas=c(george.2016$V1,george.2017$V1))
ggplot(george.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()

# Wynona Ryder
wynona.2016 <- read.table('datos/Wynona_Ryder-2016.dat')
wynona.2017 <- read.table('datos/Wynona_Ryder-2017.dat')
wynona.datos <- data.frame( dia=c(1:nrow(wynona.2016),1:nrow(wynona.2017)),
                    year=c(rep('2016',nrow(wynona.2016)),rep('2017',nrow(wynona.2017))),
                    visitas=c(wynona.2016$V1,wynona.2017$V1))
ggplot(wynona.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()

# Vaughn
vaughn.2016 <- read.table('datos/Vaughn-2016.dat')
vaughn.2017 <- read.table('datos/Vaughn-2017.dat')
vaughn.datos <- data.frame( dia=c(1:nrow(vaughn.2016),1:nrow(vaughn.2017)),
                    year=c(rep('2016',nrow(vaughn.2016)),rep('2017',nrow(vaughn.2017))),
                    visitas=c(vaughn.2016$V1,vaughn.2017$V1))
ggplot(vaughn.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()

# Vaughn
paloma.2016 <- read.table('datos/Paloma_Chamorro-2016.dat')
paloma.2017 <- read.table('datos/Paloma_Chamorro-2017.dat')
paloma.datos <- data.frame( dia=c(1:nrow(paloma.2016),1:nrow(paloma.2017)),
                    year=c(rep('2016',nrow(paloma.2016)),rep('2017',nrow(paloma.2017))),
                    visitas=c(paloma.2016$V1,paloma.2017$V1))
ggplot(paloma.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()

```

On the other hand, the graph ressembles a witch´s hat, with a noteworthy peak and a more gradual descense in the number of visits in a week´s time. Although to the date it has not returned to the original number of views, it is to be supposed that it will take place in more or less in the same period as the page we started studying, hence  seven days for the duration of the Twitter effect in the  notoriety of a public figure and the visits to its Wikipedia pages.

# Conclussion

The visits per day are multiplied by 20, more or less, the first day. The following days it loses interest, being the duration of the phenomenon approximately a week, indeed the lifetime of a meme, at least nowadays, in this country and for this type of memes, with solely a name or an identifiable person. Therefore, the data obtained for Copernico as well as similar incidents like Meryl Streep´s, seem to confirm these results that presumably will only have an universal statistic value when they are reproduced for a greater range of cases.
In addition these memes don´t tend to have much *depth*, being relegated to just a page or even to the first paragraph of it. It doesn´t seem as if the interest was spread to other concepts or people linked to it. As to the global importance of the mental occupation of the Internet´s public, it depends greatly on the initial interest of the person. The visits to the page of Meryl Streep, for instance, reach a far higher level and its position in the ranking is kept for more time, also owing to the fact that it was already between the 1000 most visited pages of the Spanish Wikipedia.

```{r meryl-cop,echo=FALSE}
meryl.cop.datos <- data.frame( dia=c(1:nrow(meryl.2017),1:nrow(datos.2017)),
                      year=c(rep('Meryl Streep',nrow(meryl.2017)),rep('Nicolás Copérnico',nrow(datos.2017))),
                    visitas=c(meryl.2017$V1,datos.2017$V1))
ggplot(meryl.cop.datos,aes(x=dia,y=visitas,group=year,col=year))+geom_line()+theme_tufte()
```

This also takes into account the fact that the *Ojeda-Copérnico incident* has awaken interest mainly in Spain, with little or none out of its politic impact. However, what happened to Meryl Streep has had  repercussion on the whole world, of course including South America, where the bigger amount of  visitors of the Spanish Wikipedia is gathered. Nevertheless, initially Meryl Streep´s visits doubled Nicolas Copernico´s, and when its notoriety started, her visits were multiplied by 25, not by 15 as in the other case.
As a conclussion, the rise in the notoriety of an incident with politic and local connotations is relatively little, compared to a cultural incident with certain politic connotations, which is global. That mindshare does not have an influence in the length of the notoriety, in both cases of about a week. 
## Note
All the data and necessary files can be found at http://github.com/JJ/Cop-rnico-visitas, and can be used with a free license. We will be very grateful given that the person that uses them references this article.

## Note

All files, data and scripts needed to generate this paper are available at http://github.com/JJ/Cop-rnico-visitas and can be used with a free license. If you use in any scientific publication we are grateful for referencing this paper or the other papers working on the same data included in the bibliography. 

## References
