---
title: "Temporal evolution of notoriety of Wikipedia pages with origin in social networks"
author: "JJ Merelo and Elena Merelo"
date: "28 de febrero de 2017"
output:
  pdf_document:
    keep_tex: true
bibliography: refs.bib
abstract:  "The Wikipedia is a web portal created by users and its
  simplicity, references and also the inclusion as insets 
  introductory paragraphs for their pages in Google search results have made it the go-to place
  to find out about current events or persons featured in them. Besides, its open application programming interface (API)
  allows any user to know about the number of visits some particular
  page has. In this paper, after certain events that made Copernicus a
  viral meme in Spain, we study the intensity and duration of the increment of visits to his
  page and other pages related to the event. Using pages related to other persons
  as a comparison, we try to establish a typical duration of notoriety achieved through social networks,
  mainly comparing with visits in previous years in the same dates. We
  conclude that a 7 day duration is the statistical mode and that this duration is relatively
  independent of the initial increase in the number of visits."
---

# Introduction

The so called “Slashdot Effect”
[@adler1999slashdot,@halavais2001slashdot], known since the end of the
90s, is the name attributed to the increase in pageviews that the
websites experience when mentioned by publications
like [Slashdot](http://slashdot.org), one of the most visited in the
area of technology and especially programming. This effect provokes a
peak in visits, especially directed to a single webpage, that in a lot
of cases and more obscure ages in which there were not any publication
of content networks, had a similar effect of a denial of service
attack. In Spain we have also referred to the “Efecto Menéame”
[@dans2008anatomia], because of the influence of this aggregator of
news and even, before and in a lesser measure, the "Efecto Barrapunto"
[@blanco2009blog] originated by mentions to the web Barrapunto.com, a
Spanish Slashdot. 

However and to the extent of our knowledge, until now there has not
been many studies analising a similar effect aroused by mentions in
social networks, maybe due to the fact that such a effect is not
caused by a single website and might be eventually directed to many
different pages; even lesser is the effect they have in the search or
mindshare of the concepts. Indeed, web analytics is currently far more
complex and generally the visits on a page, say, of a newspaper or
ecommerce site are not public. And besides, instead of studying the
effect of a single link, nowadays it is of more interest to study the
topics that pop up in the social networks. However, the effect these
have in the mindshare of the public and its behaviour while surfing
the Net has not received so much attention as the behavior of the
trending topics themselves. 

At the start of 2017, an anecdote in a Wise Men from Orient parade in
Madrid, in which a video locutor referred to a character in a float as
Colon, to which he answered “I am Copernico!” in an amusing
tone,
[as is narrated in Verne](http://verne.elpais.com/verne/2017/01/07/articulo/1483785572_720611.html),
gave way to a storm of memes, which increased on the following days
when a journalist and a politician discussed in a TV programme about
what did Copernico really do. 

In a series of articles and figures written in Spanish [merelo17, @merelo17:2, @Merelo2017:figshare]
that show the temporary evolution of this mindshare through the visits
to the Wikipedia page of Copernico [@wiki:Copernico] we have analyzed
the change in the number of visits, especially looking for the
duration of this phenomenon by means of studying the daily
number of visits. In this article, that closes the series, we try to find the causes of that duration, comparing it with
other phenomenons of visits to the Wikipedia which have happenned in a
similar time frame.

The rest of the paper is organised in the following way: first we present
the methodology that has been followed in order to capture and study
the data is presented. In the following section the visits to the
Wikipedia page of Copernico and other
related webpages is analized, along with the evolution in its
ranking of viewed pages in the Wikipedia. We will also examine the
visits to the pages of other persons that have become notorious due to
several reasons. Finally, we will draw some conclusions. 

## Methodology


The Wikipedia, through its application programming interface (API),
allows acces to the visits of each and every one of the different
pages it hosts. It is a REST interface which can be directly accessed
without any authentication. For the purposes of this paper, we tap
this API using  a script that, along with the rest of data and
programmes, is free software and hence can be downloaded from GitHub
[@Merelo2017:figshare]. The script uses `curl` for downloading the
file, just by codifying as URL the name of the page, next to
`all-agents`, and selecting as parameters the two dates in which we
are interested, from the start of 2017 until now. Due  to the fact
that the Wikipedia API returns a file in JSON, we pull the data we are
interested 
in out by means of the `jq` utility, that allows to make complex
queries to JSON files. That way, we extract the views, and we put it
in data files, which will also be attached to this paper. The rest of
the data analyzed in this paper is extracted in the same way, via
shell scripts that download an URL and the use of `jq`. 

 Nonetheless, in order to extract the ranking of the one thousand most
 visited pages, a script that gets the position in the ranking and the
 number of visits, along with the title of the article, is used. Then,
 the information is filtered, leaving only the "content" articles that are in
 the ranking, deleting special pages. Furthermore the ranking is once
 again calculated. Last, these rankings are processed to calculate the
 evolution in the position of the page as the days go by, saving in a
 file a number with the position that a specific page occupies each
 day. 
 
The results of analyzing the data are shown in the next section.


## Visits to the page of Nicolás Copérnico and other related pages

Data is processed in this paper using R; scripts can be accessed in
the source to this paper.

```{r setup,echo=FALSE}
library(ggplot2)
library(ggthemes)
datos.2016 <- read.table('datos/Nicol%C3%A1s_Cop%C3%A9rnico-2016.dat')
datos.2017 <- read.table('datos/Nicol%C3%A1s_Cop%C3%A9rnico-2017.dat')
datos <- data.frame( dia=c(1:nrow(datos.2016),1:nrow(datos.2017)),
                    year=c(rep('2016',nrow(datos.2016)),rep('2017',nrow(datos.2017))),
                    visits=c(datos.2016$V1,datos.2017$V1))
```               

First we will focus on the anecdote that originated this paper by
analyzing the visits to the Nicolás Copérnico page in the Spanish
wikipedia during 2017 and the same period in 2016. This period falls
outside the school year in Spain, with classes starting just next to
that day. That might account for a certain interest in the page, which
implies that there will be some daily variation in the number of
visits anyway. Comparing the visits for two years will help us
discount that effect. 

```{r imagen,echo=FALSE}
ggplot(datos,aes(x=dia,y=visits,group=year,col=year))+geom_line()+theme_tufte()
```
It is noticeable that, after reaching a peak the first day, it
descends the second one and even quicklier the third
one. Nevertheless, from the fourth day the fall is much softer, until
having approximately the same data a week later. In fact, there is
some variation also in the previous year, that is why we compare  the
relationship between last and this year´s visits in the next chart.

```{r ratio,echo=FALSE}
ratio <- data.frame( dia=c(1:nrow(datos.2016)),ratio=datos.2017$V1/datos.2016$V1)
ggplot(ratio,aes(x=dia,y=ratio))+geom_line()+theme_tufte()
```
From an initial situation where the number of visits is similar,
though slightly superior, it goes to having 20 times more visits,
number that slowly drops to 10 times three days past the peak. On the
fifth day the rate is of a 50% more visits; a little bit over last
year´s on the same date, taking into account that last year´s tendency
was of a gradual growth in the number of visits. Eventually, the
number of visits reaches a plateau. 

It is interesting to have a look at the evolution of some pages linked
to the same concept, especially the one of the main achievement of
Copérnico, the heliocentric theory, which motivated the following day
a debate on TV. 

```{r helio,echo=FALSE}
hc.2016 <- read.table('datos/Teor%C3%ADa_helioc%C3%A9ntrica-2016.dat')
hc.2017 <- read.table('datos/Teor%C3%ADa_helioc%C3%A9ntrica-2017.dat')
hc.datos <- data.frame( dia=c(1:nrow(hc.2016),1:nrow(hc.2017)),
                    year=c(rep('2016',nrow(hc.2016)),rep('2017',nrow(hc.2017))),
                    visits=c(hc.2016$V1,hc.2017$V1))
ggplot(hc.datos,aes(x=dia,y=visits,group=year,col=year))+geom_line()+theme_tufte()
```
The number of visits moves in a different scale and thus there are
more fluctiations along the day. However, like in the previous case,
even though initially the visits were very much alike or even
inferiors, presumably in five days it went back to its initial value,
with a small variation. In five days it has less visits than last
year, the same situation that the previous day to the incident. 

What also changes is the correlation between the number of people that
visit a web and the ones that visit the other. We have a look at it in the next graphic.

```{r ratiohc,echo=FALSE}
ratio.hc.2016 <- datos.2016/hc.2016
ratio.hc.2017 <- datos.2017/hc.2017
hc.ratios <- data.frame( dia=c(1:nrow(ratio.hc.2016),1:nrow(ratio.hc.2017)),
                    year=c(rep('2016',nrow(ratio.hc.2016)),rep('2017',nrow(ratio.hc.2017))),
                    visits=c(ratio.hc.2016$V1,ratio.hc.2017$V1))
ggplot(hc.ratios,aes(x=dia,y=visits,group=year,col=year))+geom_line()+theme_tufte()
```

In the year 2016 there is a rate inferior to 2, consequently the page
of the heliocentric theory receives more or less half the visits that
Copérnico´s, with little variations. We suppose that there is more
interest in the character than in its contribution, though not a
lot. On the other hand, when the issue “Copérnico-Ojeda" breaks out,
all of a sudden the people get curious about the character, without
caring much about his contribution to the scientific world. It is to
important to point out that we do not know whether they are the same
people that visit the webpage, because of the fact that the path they
follow in Wikipedia is not returned by the API. Anyway, it is reasonable
to suppose that a big percentage of the people that visit
whichever of the pages will end up following the links in order to
visit the other one. The link to the heliocentric theory appears in
the first lines of Copérnico´s page and it is the thirteenth link from
the start of the page, while Copérnico shows up int the second
paragraph of the theory page, being the tenth link in this case.

Let´s also see another page, Galileo Galilei´s, who spread the
heliocentric theory. His name  does not show up in any way in the
incident but to check the visits here would be relevant to study the interest these kind of
pages have and whether there is a offshoot of interest from one to
another. 

```{r galileo,echo=FALSE}
galileo.2016 <- read.table('datos/2016-galileo.dat')
galileo.2017 <- read.table('datos/2017-galileo.dat')
galileo.datos <- data.frame( dia=c(1:nrow(galileo.2016),1:nrow(galileo.2017)),
                    year=c(rep('2016',nrow(galileo.2016)),rep('2017',nrow(galileo.2017))),
                    visits=c(galileo.2016$V1,galileo.2017$V1))
ggplot(galileo.datos,aes(x=dia,y=visits,group=year,col=year))+geom_line()+theme_tufte()
```

Apart from the curiosity on  do they evolve in the same way, with a
considerable drop the seventh day, there is neither an obvious
tendency nor can we suppose that pages of this kind of content could
have an increased interest. In fact, the twelfth day there were less
visits in Galileo´s page than last year. Although there is a chance of
a transfer of attention and hence of visits from one to another, since
Galileo´s is linked to Copérnicos, and as there aren´t changes in the
same way as in Copérnico´s rules out a sudden interest in the
Renaissance Science and the heliocentric theory like a cause of the
rise in visits that will be then attributable to the incident. 

From this analysis we can conclude that the interest in a particular
Wikipedia page famous through a meme in the social networks lasts
around seven days, and that it is a shallow interest extending
exclusively to the page itself, not any concepts related to it. Let us
see next what happens to other similar pages.

## Duration of notability in other Wikipedia pages. 

We also examine the visits to the page of Meryl Streep, whose speech during the Golden Globes had a considerable impact, in the same way as the elected president´s reaction did. In this case it is an example of a page that on its own has a lot of visits. We plot it side by side the visits to the page of John Hurt, who died on January 21st, 2017. 

```{r meryl,echo=FALSE,fig.width=4,fig.align='top'}
meryl.2016 <- read.table('datos/Meryl_Streep-2016.dat')
meryl.2017 <- read.table('datos/Meryl_Streep-2017.dat')
john.2016 <- read.table('datos/John_Hurt-2016.dat')
john.2017 <- read.table('datos/John_Hurt-2017.dat')
meryl.datos <- data.frame( dia=c(1:nrow(meryl.2016),1:nrow(meryl.2017)),
                    year=c(rep('2016',nrow(meryl.2016)),rep('2017',nrow(meryl.2017))),
                    visits=c(meryl.2016$V1,meryl.2017$V1))
ggplot(meryl.datos,aes(x=dia,y=visits,group=year,col=year))+geom_point()+geom_smooth(span=0.25)+theme_tufte()

john.datos <- data.frame( dia=c(1:nrow(john.2016),1:nrow(john.2017)),
                    year=c(rep('2016',nrow(john.2016)),rep('2017',nrow(john.2017))),
                    visits=c(john.2016$V1,john.2017$V1))
ggplot(john.datos,aes(x=dia,y=visits,group=year,col=year))+geom_point()+geom_smooth(span=0.25)+theme_tufte()
```

The number of visits reached by John Hurt's wikipedia page is almost 3 times as high as the one reached by Meryl Streep. The mechanism at work here must go in the opposite direction of fame: John Hurt was not probably known by name, so people looked him up much more than Meryl Streep, who is a household name and whose movies are shown continously, even in Spanish TV channels. But even with the different scales, the *relaxation time* hovers around 7 days, a few more in the case of Meryl Streep (left). Besides, when the visits plateau the *new normal* is slightly superior to the visits achieved in the previous year or before the incident that initiated the sudden fame. Let us see this in the second case, the John Hurt page.

```{r ratiojohn,echo=FALSE}
ratio <- data.frame( day=c(1:nrow(john.2016)),ratio=john.2017$V1/john.2016$V1)
ggplot(ratio,aes(x=day,y=ratio,color='Ratio'))+geom_line()+scale_y_log10() + theme_tufte() + geom_hline(aes(yintercept=1))
```

We plot the *y* axis in a logarithmic scale to see more clearly the ration after *relaxation*. The straight line goes through the ratio == 1. Although there are variations, for a few days after the peak ratio is still slightly higher than 1, which seems to imply that the *mindshare* of this particular meme follows two different phases: one of sharp decline, that lasts the aforementioned 7-10 days, followed by a plateau where the decline is much slower, and stais stationary with a ratio higher than one, in the same way as was seen before.

These tho regimes, duration and notability need not take the same time, and will probably vary along time. We have also examined the cases of three other memes: from left to right, the death of Robert Vaughn in November 2016 and Carrie Fisher and George Michael in December 2016. Our intention was, looking at these events taking place a few months before, if there is any variation. 

```{r carrie,echo=FALSE,fig.width=2.75, fig.height=4,fig.align='top'}
# Vaughn
vaughn.2016 <- read.table('datos/Vaughn-2015.dat')
vaughn.2017 <- read.table('datos/Vaughn-2016.dat')
vaughn.datos <- data.frame( dia=c(1:nrow(vaughn.2016),1:nrow(vaughn.2017)),
                    year=c(rep('2015',nrow(vaughn.2016)),rep('2016',nrow(vaughn.2017))),
                    visits=c(vaughn.2016$V1,vaughn.2017$V1))
ggplot(vaughn.datos,aes(x=dia,y=visits,group=year,col=year))+geom_point()+geom_smooth(span=0.25)+theme_tufte()

carrie.2016 <- read.table('datos/Carrie-Fisher-2016.dat')
carrie.2017 <- read.table('datos/Carrie-Fisher-2017.dat')
carrie.datos <- data.frame( dia=c(1:nrow(carrie.2016),1:nrow(carrie.2017)),
                    year=c(rep('2016',nrow(carrie.2016)),rep('2017',nrow(carrie.2017))),
                    visits=c(carrie.2016$V1,carrie.2017$V1))
ggplot(carrie.datos,aes(x=dia,y=visits,group=year,col=year))+geom_point()+geom_smooth(span=0.25)+theme_tufte()

# George Michael
george.2016 <- read.table('datos/George-Michael-2016.dat')
george.2017 <- read.table('datos/George-Michael-2017.dat')
george.datos <- data.frame( dia=c(1:nrow(george.2016),1:nrow(george.2017)),
                    year=c(rep('2016',nrow(george.2016)),rep('2017',nrow(george.2017))),
                    visits=c(george.2016$V1,george.2017$V1))
ggplot(george.datos,aes(x=dia,y=visits,group=year,col=year))+geom_point()+geom_smooth(span=0.25)+theme_tufte()
```

Let us first look at the first chart on the left, representing visits of the Robert Vaughn page. Being not so well known in pop culture, the peak reached is relatively small, but still visits multiplied 100-fold when his death was known. Carrie Fisher (center) and George Michael (right), being better known, had a higher level of visits, but the visits reached multiply 200-fold, and in this case the two regimes can be observed: sharp decrease that lasts 10 days followd by a plateau where the ratio to visits before the event  is around 2-3. The same type of behavious can be observed about the visits to the page of George Michael. 

Finally, let us look at similar events that happened later along the time line: the viral sensation of Winona Ryder showing a range of facial expressions during a gala (left), and the death of a journalist and famous 80s musical TV anchor in Spain, Paloma Chamorro (right).

```{r winona,echo=FALSE,fig.width=4,fig.align='top'}
# Winona Ryder
winona.2016 <- read.table('datos/Winona_Ryder-2016.dat')
winona.2017 <- read.table('datos/Winona_Ryder-2017.dat')
winona.datos <- data.frame( dia=c(1:nrow(winona.2016),1:nrow(winona.2017)),
                    year=c(rep('2016',nrow(winona.2016)),rep('2017',nrow(winona.2017))),
                    visits=c(winona.2016$V1,winona.2017$V1))
ggplot(winona.datos,aes(x=dia,y=visits,group=year,col=year))+geom_line()+theme_tufte()


# Paloma Chamorro
paloma.2016 <- read.table('datos/Paloma_Chamorro-2016.dat')
paloma.2017 <- read.table('datos/Paloma_Chamorro-2017.dat')
paloma.datos <- data.frame( dia=c(1:nrow(paloma.2016),1:nrow(paloma.2017)),
                    year=c(rep('2016',nrow(paloma.2016)),rep('2017',nrow(paloma.2017))),
                    visits=c(paloma.2016$V1,paloma.2017$V1))
ggplot(paloma.datos,aes(x=dia,y=visits,group=year,col=year))+geom_line()+theme_tufte()

```
The chart on the left, although with some *noise*, shows more or less the same phenomenon of a sharp peak followed by rapid decrease in visits, that take between 7 and 10 days. The lower number of visits reached probably accounts for the fact that the meme was self-contained (a series of faces) and did not need additional information on the person. However, the one on the right is entirely familiar in scale and size, with *rebounds* probably due to published obituaries and settling to a number of visits that is slightly superior than originally. 

Eventually, all visit graphs ressemble a "witch´s hat"", with a noteworthy peak and a more gradual decrease in the number of visits in approximately a week´s time, although usually a bit longer. We will examine this findings in the next section, where we present our conclussions. 

## Conclussion

By examining the visits of Wikipedia pages through its open API we intend to look for patterns in these visits and, through them, establish the duration of Internet memes, assuming that when one particular person or concept becomes noticeable, people will check out their Wikipedia page as first source of information.

What we have found by looking at nine different pages corresponding to 9 different persons that have become trending topics in social networks, but specially in Spain, starting with a very local event that actually initiated this research. What we have found analyzing visit data is that noteworthiness lasts for at least a week, and sometimes a bit more, up to 10 days; from then on, visits go back to the usual rate but at a much slower pace. The headline would thus be "Fame in social networks lasts for only a week", but this would have some caveats that would be worth checking out in the future, such as the relationship with the nature of the meme itself (due to an unexpected event in the life of the person, for instance, or some other thing). A secondary finding is that the magnitude of the increase in visits is related to prior fame: "those who have, will get more", reaching a much higher ration between prior and post-event visits. 

In general, the use of this API by Wikipedia opens a whole world of possibilities for the research of collective behavior. One of them would be to analyze what part of these visits were done using different devices, and which ones correspond to edits or simple visits. The difference between the increment of visits of life-changing events or other kind of memes might be due to this fact. This could be an interesting venue of research in the near future, as well as checking if there is in fact some change in this one-week duration of memes. 


## Note

All files, data and scripts needed to generate this paper are available at http://github.com/JJ/Cop-rnico-visitas and can be used with a free license. If you use in any scientific publication we are grateful for referencing this paper or the other papers working on the same data included in the bibliography. 

## References
